A Shared Subspace Model for Multi-Group Covariance Estimation

### 2016-06-30 paper/mscov.tex
Latest iteration.

\title{Shared Subspace Model}

\date{\today}

\title{Shared Subspace Covariance Model}
\author{Alexander Franks and Peter Hoff}
\date{\today}
\maketitle

@@

  We develop a model-based method for evaluating similarities and
  differences among several $p\times p$ covariance matrices in the
  $p \gg n$ setting.  This is done by assuming a spiked covariance
  model for each group and sharing information about the space spanned
  by the group-level eigenvectors.  We use empirical Bayes to identify
  a low-dimensional subspace which explains as a large amount of
  variation across all groups, and, given this subspace, show how a
  Gibbs sampling algorithm can be used to estimate the posterior
  uncertainty of group-level eigenvectors and eigenvalues.  The
  implementation and utility of our model is illustrated with analyses
  of high-dimensional multivariate biological data.

@@

In many natural systems, high dimensional data is often very
structured and thus can be well understood on a relatively low
dimensional subspace. For example, with gene expression data, the
effective dimensionality is thought to scale with the number of gene
regulatory modules, not the number of genes themselves
\citep{Heimberg2016}.  As such, differences in gene expression across
groups should be expressed in terms of differences between these
regulatory modules rather than strict differences between expression
levels.  Such differences can be examined on a subspace that reflects
the correlations resulting from these modules.

It has long been known that the sample covariance matrix is a poor
estimator of the population covariance matrix, particularly when the
sample size is comparable to or smaller than the number of features
($n \leq p$) \citep{Dempster1969, Stein1975}.  A host of estimators
have been devised to address this fact. These primarily include
shrinkage approaches \citep{Schafer2005, Ledoit2011} and low
rank factor models \citep{Mardia1980, Ullman2003, Fan2008}.  In many
cases, a covariance matrix is highly structured and thus can be
sparsely parameterized even if the matrix itself is not low rank
\citep{Burg1982, Williams1993} .  Along these lines, there has been a
recent interest in sparse Gaussian graphical model, in which the there
are assumed to be many zeros in the inverse covariance matrix
\citep{Meinshausen2006, Friedman2008}.

Most of these approaches assume that the goal is to estimate the
covariance matrix of a single population.  In many applications,
however, the data can be split into groups, each of which represent
samples from populations with distinct (but possibly related)
covariance matrices.  In this case, a comparative analysis of
covariance patterns across multiple groups of observations is often
warranted.  Evaluating similarities and differences between covariance
matrices can be an important complement to classical analyses which
center primarily on estimating differences in means across groups.
For instance, in case control-studies, mean level effects may be small
relative to subject variability; distributional differences between
groups may still be evident as differences in the covariances between
features.  Even when mean level differences are detectable, better
estimates of the covariability of features across groups may lead to
an improved understanding of the mechanisms underlying these apparent
mean level differences.  Further, accurate covariance estimation is an
essential part of many prediction tasks (e.g. quadratic discriminant
analysis).

Some of the earliest approaches for modeling multiple
covariance matrices focused on estimation in terms of the spectral
decompositions for these matrices.  \cite{Flury1987} developed
estimation and testing procedures for the ``common principal
components'' model, in which a set of covariance matrices were assumed
to share the same eigenvectors.  \citet{Schott1991, Schott1999}
considered cases in which only certain eigenvectors are shared across
populations, and \citet{Boik2002} described an even more general model
in which eigenvectors can be shared between all or some of the groups.
More recently, \citet{Hoff2009}, noting that eigenvectors are unlikely
to be shared exactly between groups, introduced a hierarchical model
for eigenvector shrinkage based on the matrix Bingham distribution.
There has also been a significant interest in estimating covariance
matrices using Gaussian graphical models. For Gaussian graphical
models, zeros in the precision matrix correspond to conditional
independences between pairs of features given the remaining features
\citep{Meinshausen2006}.  \citet{Witten2014} extended existing work in
this area to the multi-group setting, by pooling information about the
pattern of zeros across precision matrices.

@@

Another popular method for modeling relationships between
high-dimensional multivariate data is partial least squares regression
(PLS). This approach, which is a special case of a bilinear factor
model, involves projecting the data into a lower dimension space which
maximizes the similarity of the two groups.  This technique does not
require the data from each group to share the same feature set.  An
common variant for prediction, partial least squares discriminant
analysis (PLS-DA) is especially common in chemometrics and
bioinformatics \citep{Barker2003} .  Although closely related to the
approaches we will consider here, the primarily goal of of PLS-based
models is to create regression or discrimination models, not to
explicitly infer covariance matrices from multiple groups of data.
Nevertheless, the basic idea that data can often be well represented
on a low dimensional space is an appealing one that we leverage as
well.

In this paper we propose a multi-group covariance estimation model by
sharing information about the subspace spanned by group-level
eigenvectors.  The shared subspace assumption can be used to improve
estimates and facilitate the interpretation of differences between
covariance matrices across groups.  For each group, we assume ``the
spiked covariance model'', a well studied variant of the factor model
\citep{Johnstone2001}.  In Section \ref{sec:shared} we briefly review
the behavior of spiked covariance models for estimating a single
covariance matrix and then introduce our extension to the multi-group
setting.  In Section \ref{sec:inference} we describe an optimization
algorithm for inferring the shared subspace and then outline a MCMC
algorithm for estimating the posterior distribution of the projected
covariance matrices.  In Section \ref{sec:simulation} we
investigate the behavior of these models in simulation and then
demonstrate the utility of the model with real biological data in
Section \ref{sec:app}.

@@

Classical results for parametric models (e.g., \citet{Schwartz1965})
imply that asymptotically in $n$ for fixed $p$, an estimator will
be consistent for a spiked population covariance as long as the
assumed number of spikes (eigenvalues larger than $\sigma^2$) is greater than or
equal to the true number.  However, when $p$ is large relative to $n$,
as is the case for the examples considered here, things are more
difficult.  Under the spiked covariance model, it has been shown that
if $p/n \rightarrow \gamma >0$ as $n\rightarrow \infty$, the $k$th largest
eigenvalue of $\bl S/(n\sigma^2)$ will converge to an upwardly biased
version of $\lambda_{k}+1$, if $\lambda_k$
is greater than $\sqrt{\gamma}$ \citep{Baik2006, Paul2007} .  This has
led several authors to suggest estimating $\Sigma$ via shrinkage of
the eigenvalues of the sample covariance matrix. In particular, in the
setting that $\sigma^2$ is known, \citet{Donoho2013} propose estimating
all eigenvalues whose sample estimates are smaller than 
$\sigma^2(1+\sqrt{\gamma})^2$ by $\sigma^2$, and shrinking the
larger eigenvalues in a way that depends on the particular loss
function being used.  These shrinkage functions are shown to be
asymptotically optimal in the $p/n\rightarrow \gamma$ setting.

@@

When estimating a single covariance matrix, the absence of prior
knowledge suggests estimators that are equivariant with respect to
rotations and scale changes.  The situation is different when we are
interested in estimating multiple covariance matrices from distinct,
but related groups.  Here, group-level equivariance to rotations is an
unreasonable assumption; both eigenvalue \emph{and} eigenvector
shrinkage can play an important role in improving covariance
estimates.

For multi-group covariance estimation, we assume that we have $K$
matrices, $Y_1, ... Y_K$, where $Y_k$ conists of where $Y_k$ consists
of $n_k$ independent rows of mean-zero normal data of dimension $p$,
typically with $n_k \ll p$.  Then, $Y_k^TY_k = \bl S_k$ has a
(degenerate) Wishart distribution as in Equation \ref{eqn:lik}.  To
improve estimation, we seek estimators for the population covariance
from one group that depend on the data from all $K$ groups, e.g. of
the form $\hat \Sigma_k(\bl S_1, ... \bl S_K)$.  To this end, we
extend the spiked covariance model to grouped data, by assuming that
the anisotropic variability from each group occurs on a common low
dimensional subspace.  Specifically, we posit that the covariance
matrix for each group can be written as

\begin{equation}
\Sigma_k = V\Psi_kV^T + \sigma^2_kI.
\label{eqn:sspsi}
\end{equation}
%
\noindent Here, the columns of $V \in \mathcal{V}_{p, s}$ determine a subspace
shared by all groups $k$ with $p \geq s$.  Throughout this paper we
will denote to the shared subspace as $VV^T \in \mathcal G_{r,p}$, where
$\mathcal G_{r,p}$, known as the Grassmann manifold, consists of all
$s$-dimensional linear subspaces of $\mathcal{R}^p$
\citep{Chikuse2012}.  Although $V$ is only identifiable up to right
rotations, the matrix $VV^T$, which defines the plane of variation shared by all
groups, is identifiable.  Later, to emphasize the connection to the
spiked PCA model (\ref{eqn:spiked}), we will write $\Psi_k$ in terms
of its eigendecomposition, $\Psi_k = O_k\Lambda_kO_k$, where $O_k$ are
eigenvectors and $\Lambda_k$ are the eigenvalues of $\Psi_k$ (see
Section \ref{sec:bayes}).

@@

In this section we outline our empirical Bayesian approach for estimating
covariance parameters and their associated uncertainty.  
Uncertainty quantification is particularly important in the $n \ll p$
setting where seemingly large differences in the point estimates of
the low rank covariance matrices across groups may not actually
reflect significant differences.  Although uncertainty quantification
for all parameters is desirable, for the size of the problems we are
generally interested in, we found MCMC inference for the subspace
$VV^T$ to be intractable.  Our optimization based approached is
significantly faster and works well for high-dimensional datasets.  As
such, we start, in Section \ref{sec:em}, by outlining an
expectation-maximization algorithm for estimating the shared
subspace. In Section \ref{sec:bayes} we describe a Gibbs sampler for
posterior inference of $\Psi_k$ given the common subspace.  These two
algorithms can be used in conjunction to first identify and evaluate a
shared subspace and then to quantify the uncertainty associated with
the projected covariance matrices.

TODO: emphasize 1) justification 2) objective of \emph{identifying} this
subspace when p much greater than n.

@@

We propose an expectation-maximization (EM) algorithm, to estimate the
relevant quantities.  In particular we maximize the marginal posterior
distribution of $V$, treating $M_k^{-1}$ and $\frac{1}{\sigma_k^2}$ as
the ``missing'' parameters to be imputed.  We assume independent
Jeffreys' prior distributions for both $\frac{1}{\sigma_k^2}$ and
$M_k^{-1}$ and a uniform prior for $V \in \mathcal{V}_{s, p}$.
Jeffreys' prior for these quantities corresponds to
$p(\frac{1}{\sigma_k^2}) \propto \sigma_k^2$ and
$p(M_k^{-1}) \propto |M_k^{-1}|^{-(s+1)/2}$.  From the likelihood it
can easily be shown that the conditional posterior for $M_k^{-1}$ is

@@

\noindent which is a gamma($n_k(p-r)/2$,
$\frac{1}{2}tr((I-VV^T)\mathbf{S}_k)$) distribution.  Based on these
results the following EM algorithm can be used.

@@

For the ``M-step'', we use a numerical optimization algorithm based on
the Cayley transform to preserve the orthogonality constraints in $V$
\citep{Wen2013}.  Importantly, the complexity of this algorithm is
dominated by dimension of the shared subspace, $s$, not the number of
features $p$.  Thus, our approach is computationally efficient for
relatively small values of $s$, even when $p$ is large.

@@

It is important to quantify how much of the variability in each group can be
explained by an inferred subspace.  If $V$ is a basis for
the shared subspace, then for each group $k$, most of the variation in
$Y_k$ should be preserved when projecting the data to this subspace.
To characterize the extent to which this is true for different groups, we
propose a simple estimator for the proportion of ``signal'' variance
that lies on any given subspace.  Specifically, for an estimated
subspace $\hat{V}$, we apply the following statistic for the ratio of the
sum of the first $s$ eigenvalues of $\hat{V}^T\bl \Sigma_k \hat{V}$ to
the sum of the first $s$ eigenvalues of $\Sigma_k$:
%
\begin{equation}
r(Y_k: \hat{V}, \hat{\sigma}_k^2) = \frac{||Y_k\hat{V}||_F}{\underset{\widetilde{V} \in \mathcal{V}_{p, s}}{\text{max}}
  ||Y_k\widetilde{V}||_F - \hat{\sigma}_k^2ps}
\label{eqn:ratio}
\end{equation}
%
\noindent where $||.||_F$ is the Frobenius norm. Our motivation for
this statisic is the fact that if
$\Sigma_k = \hat{V}\Psi_k\hat{V}^T + \sigma_k^2I$ then
$r(Y_k: \hat{V}, \hat{\sigma}_k^2) \rightarrow 1$.  To see this, note
that $\underset{V \in \mathcal{V}_{p, s}}{max}\frac{||Y_kV||_F}{n_k}$
is equivalent to the sum of the first $s$ eigenvalues of the sample
covariance matrix $\bl S_k/n_k$.  With $\hat{\lambda}^{(k)}_i$
the $i$-th eigenvalue of $\frac{\bl S_k}{n_k}$, \citet{Baik2006}
demonstrated that asymptotically as $p, n_k \rightarrow \infty$ with
$\frac{p}{n_k} = \gamma_k$ fixed

\begin{eqnarray}
\hat{\lambda}^{(k)}_i &\rightarrow& \lambda^{(k)}_i\left(1 +
                                    \frac{\sigma_k^2\gamma}{\lambda^{(k)}_i
                                    - 1}\right)\\
& \approx& \lambda^{(k)} + \sigma^2_k\frac{p}{n_k}
\end{eqnarray}

Thus,
$\sum_i^s \hat{\lambda}^{(k)}_i \approx \sum_i^s \lambda^{(k)}_i +
\sigma_k^2ps$.
If we assume that $s$ is fixed in this asymptotic regime, then
$||Y_kV||_F/n_k$ is a consistent estimator for
$\sum_i^s \lambda^{(k)}_i$.  As a consequence, the proposed estimator
will be close to one for all groups $k$ when $\hat{V}\hat{V}^T = VV^T$
and smaller if not.  The metric can provides a useful indicator for
which groups can be reasonably compared on a given subspace and which
groups cannot.  When $r(Y_k: \hat{V}, \hat{\sigma})k^2)$ is small for
some groups, it may suggest that the rank, $s$, of the inferred
subspace needs to be larger to capture the variation in all
groups. The use of this statistic is illustrated in Section
\ref{sec:simulation}, where we plot these estimates for different
simulated datasets, and in Section \ref{sec:app}, where we plot the
estimates for subspaces inferred with real biological data.

@@

Although the EM algorithm presented in the previous section yields
point estimates for $VV^T$ and $\Psi_k$, it does not lead to natural
uncertainty for these estimates. As such, in this section, we assume
that the subspace $VV^T$ is fixed and known and focus on uncertainty
quantification for $\Psi_k$ only.  Note that when the subspace is
known, the posterior distribution of $\Sigma_k$ is conditionally
independent from the other groups, so that we can independently
estimate the conditional posterior distribution for each group.

There are many different prior specifications for $\Psi_k$ and
$\sigma^2_k$ that we could consider for estimating the posterior
uncertaity.  In this paper, we build on recent interest in the spiked
covariance model \citep{Donoho2013, Paul2007} and develop tractable
models for the eigendecomposition of $\Psi_k$.  Such models are
especially appealing because it is simple to summarize and compare
$\Psi_k$ across groups in terms their eigenstructure (See Section
\ref{sec:simulation} and \ref{sec:app}).

By modeling the eigenstructure, we can now view each covariance
$\Sigma_k$ in terms of the spiked principle components model (\ref{eqn:spiked}).
Equation \ref{eqn:sspsi}, written as a function $V$, becomes

\begin{align}
\nonumber \Psi_k &= O_k\Lambda_KO_k^T\\
\Sigma_k &= VO_k\Lambda_KO_k^TV^T + \sigma^2_kI
\label{eqn:ss}
\end{align}
%
\noindent Here, we make one further generalization to allow
$\Psi_k$ to be a rank $r \leq s$ dimensional covariance
matrix on the $s$-dimensional subspace.  Thus, $\Lambda_k$ is an $r \times
r$ diagonal matrix of eigenvalues, and $O_k \in
\mathcal{V}_{s,r}$ is the matrix of eigenvectors of
$\Psi_k$.  For any individual group, this corresponds to the original spiked
PCA model (Equation \ref{eqn:spiked}) with $U_k = VO_k \in
\mathcal{V}_{p, r}$.    This ``double spiked'' covariance model TODO

@@

\paragraph{Conditional distribution for $\Omega$:} Here we specify the conditional
distribution of the diagonal matrix $\Omega_k =
\Lambda_k(I+\Lambda_k)^{-1} = \text{diag}(\omega_{k1}, ... \omega_{kr})$.  We first consider a uniform(0,1) prior distribution for each element of $\Omega$, or
equivalently, an $F_{2,2}$ prior distribution for the elements of
$\Lambda$.  The full conditional distribution of an
element $\omega_i$ of $\Omega$ is proportional to the likelihood
function, so

@@

\subsection{Validation of Inference}

In this section we start by validating, in simulation, that our
inferential algorithm recovers the true parameters of the data
generating mechanism.  We simulate five groups of $p=200$ dimensional
data each with sample size $n_k=50$.  We choose the eigenvectors of
the anisotropic component of the covariance matrix, $\Psi_k$, to share
a common two dimensional subspace.  To fit the model to this simulated
data, we first estimate the shared subspace using the EM algorithm outlined
in Section \ref{sec:em} and then use the MCMC algorithm described in
Section \ref{sec:bayes} to estimate the posterior distribution of the
projected covariance matrices $\Psi_k$ and noise variance $\sigma_k^2$.

>>>>

\title{A Shared Subspace Model for Multi-Group Covariance Estimation}
\author{Alexander Franks and Peter Hoff}
\date{\today}
\maketitle

@@

  We develop a model-based method for evaluating similarities and
  differences among several $p\times p$ covariance matrices in the
  $p \gg n$ setting.  This is done by assuming a spiked covariance
  model for each group and sharing information about the space spanned
  by the group-level eigenvectors.  We use empirical Bayes to identify
  a low-dimensional subspace which explains variation across all
  groups and use an MCMC algorithm to estimate the posterior
  uncertainty of eigenvectors and eigenvalues on this subspace.  The
  implementation and utility of our model is illustrated with analyses
  of high-dimensional multivariate gene-expression data.

@@

% For example, it has long been known that the sample covariance matrix
% is a poor estimator of the population covariance matrix, particularly
% when the sample size is comparable to or smaller than the number of
% features ($n \leq p$) \citep{Dempster1969, Stein1975}.  A host of
% estimators have been devised to address this fact. These primarily
% include shrinkage approaches \citep{Schafer2005, Ledoit2011} and low
% rank factor models \citep{Mardia1980, Ullman2003, Fan2008}.  In many
% cases, a covariance matrix is highly structured and thus can be
% sparsely parameterized even if the matrix itself is not low rank
% \citep{Burg1982, Williams1993} .  Along these lines, there has been a
% recent interest in sparse Gaussian graphical model, in which the there
% are assumed to be many zeros in the inverse covariance matrix
% \citep{Meinshausen2006, Friedman2008}.

In modern analyses, multivariate data is often partitioned into
groups, each of which represent samples from populations with
distinct, but possibly related, distributions.  Although historically the
primary focus has been on identifying mean-level differences between
populations, there has been a growing need to identify differences in
population covariances as well.  For instance, in case
control-studies, mean level effects may be small relative to subject
variability; distributional differences between groups may still be
evident as differences in the covariances between features.  Even when
mean level differences are detectable, better estimates of the
covariability of features across groups may lead to an improved
understanding of the mechanisms underlying these apparent mean level
differences.  Further, accurate covariance estimation is an essential
part of many prediction tasks (e.g. quadratic discriminant analysis).
Thus, evaluating similarities and differences between covariance
matrices can be an important complement to more traditional analyses
for estimating differences in means across
groups.

To address this needs, we develop a novel method for
group-covariance estimation.  In our method, we exploit the fact that
in many natural systems, high dimensional data is often very
structured and thus can be best understood on a lower dimensional
subspace. For example, with gene expression data, the effective
dimensionality is thought to scale with the number of gene regulatory
modules, not the number of genes themselves \citep{Heimberg2016}.  As
such, differences in gene expression across groups should be expressed
in terms of differences between these regulatory modules rather than
strict differences between expression levels.  Such differences can be
examined on a subspace that reflects the correlations resulting from
these modules.  In contrast to most existing approaches for group
covariance estimation, our approach is to directly infer such
subspaces from groups of related data.

% Existing approaches incorporate a variety of
% different structural assumptions about the relationship between
% multiple groups of covariance matrices.

Some of the earliest approaches for multi-group covariance estimation
focus on estimation in terms of spectral decompositions.
\cite{Flury1987} developed estimation and testing procedures for the
``common principal components'' model, in which a set of covariance
matrices were assumed to share the same eigenvectors.
\citet{Schott1991, Schott1999} considered cases in which only certain
eigenvectors are shared across populations, and \citet{Boik2002}
described an even more general model in which eigenvectors can be
shared between some or all of the groups.  More recently,
\citet{Hoff2009}, noting that eigenvectors are unlikely to be shared
exactly between groups, introduced a hierarchical model for
eigenvector shrinkage based on the matrix Bingham distribution.  There
has also been a significant interest in estimating covariance matrices
using Gaussian graphical models. For Gaussian graphical models, zeros
in the precision matrix correspond to conditional independences
between pairs of features given the remaining features
\citep{Meinshausen2006}.  \citet{Witten2014} extended existing work in
this area to the multi-group setting, by pooling information about the
pattern of zeros across precision matrices.

@@

Another popular method for modeling relationships between
high-dimensional multivariate data is partial least squares regression
(PLS). This approach, which is a special case of a bilinear factor
model, involves projecting the data into a lower dimension space which
maximizes the similarity of the two groups.  This technique does not
require the data from each group to share the same feature set.  A
common variant for prediction, partial least squares discriminant
analysis (PLS-DA) is especially common in chemometrics and
bioinformatics \citep{Barker2003} .  Although closely related to the
approaches we will consider here, the primarily goal of of PLS-based
models is to create regression or discrimination models, not to
explicitly infer covariance matrices from multiple groups of data.
Nevertheless, the basic idea that data can often be well represented
on a low dimensional space is an appealing one that we leverage.

In this paper we propose a multi-group covariance estimation model by
sharing information about the subspace spanned by group-level
eigenvectors.  The shared subspace assumption can be used to improve
estimates and facilitate the interpretation of differences between
covariance matrices across groups.  For each group, we assume ``the
spiked covariance model'', a well studied variant of the factor model
\citep{Johnstone2001}.  In Section \ref{sec:shared} we briefly review
the behavior of spiked covariance models for estimating a single
covariance matrix and then introduce our extension to the multi-group
setting.  In Section \ref{sec:inference} we describe an empirical
Bayes algorithm for inferring the shared subspace and estimating the
posterior distribution of the covariance matrices of the data
projected onto this subspace.  In Section \ref{sec:simulation} we
investigate the behavior of these models in simulation and demonstrate
how the shared subspace is widely applicable, even when there is
little similarity in the covariance matrices across groups.  We then
demonstrate the utility of the model in an analysis of gene expression
data from juvenile leukemia patients in Section \ref{sec:app}.  Despite the
large feature size ($p > 3000$) and small sample size ($n < 100$ per
group), we identify interpretable similarities and differences in gene
covariances on a low dimensional subspace.

@@

Classical results for parametric models (e.g., \citet{Schwartz1965})
imply that asymptotically in $n$ for fixed $p$, an estimator will be
consistent for a spiked population covariance as long as the assumed
number of spikes (eigenvalues larger than $\sigma^2$) is greater than
or equal to the true number.  However, when $p$ is large relative to
$n$, as is the case for the examples considered here, things are more
difficult.  Under the spiked covariance model, it has been shown that
if $p/n \rightarrow \alpha >0$ as $n\rightarrow \infty$, the $k$th
largest eigenvalue of $\bl S/(n\sigma^2)$ will converge to an upwardly
biased version of $\lambda_{k}+1$, if $\lambda_k$ is greater than
$\sqrt{\alpha}$ \citep{Baik2006, Paul2007} .  This has led several
authors to suggest estimating $\Sigma$ via shrinkage of the
eigenvalues of the sample covariance matrix. In particular, in the
setting that $\sigma^2$ is known, \citet{Donoho2013} propose
estimating all eigenvalues whose sample estimates are smaller than
$\sigma^2(1+\sqrt{\alpha})^2$ by $\sigma^2$, and shrinking the larger
eigenvalues in a way that depends on the particular loss function
being used.  These shrinkage functions are shown to be asymptotically
optimal in the $p/n\rightarrow \alpha$ setting.  Note that covariance
estimators of this form are are equivariant with respect to rotations
and scale changes.  The situation should be different, however, when
we are interested in estimating multiple covariance matrices from
distinct, but related groups.  Here, group-level equivariance to
rotations is an unreasonable assumption; both eigenvalue \emph{and}
eigenvector shrinkage can play an important role in improving
covariance estimates.

@@

For multi-group covariance estimation, we assume that we have $K$
matrices, $Y_1, ... Y_K$, where $Y_k$ consists
of $n_k$ independent rows of mean-zero normal data of dimension $p$,
typically with $n_k \ll p$.  Then, $Y_k^TY_k = \bl S_k$ has a
(degenerate) Wishart distribution as in Equation \ref{eqn:lik}.  To
improve estimation, we seek estimators for the population covariance
from one group that depend on the data from all $K$ groups, e.g. of
the form $\hat \Sigma_k(\bl S_1, ... \bl S_K)$.  To this end, we
extend the spiked covariance model to grouped data, by assuming that
the anisotropic variability from each group occurs on a common low
dimensional subspace.  Specifically, we posit that the covariance
matrix for each group can be written as

\begin{equation}
\Sigma_k = V\Psi_kV^T + \sigma^2_kI.
\label{eqn:sspsi}
\end{equation}
%
\noindent Here, the columns of $V \in \mathcal{V}_{p, s}$ determine a subspace
shared by all groups $k$ with $s \geq p$.  Throughout this paper we
will denote to the shared subspace as $VV^T \in \mathcal G_{p,s}$, where
$\mathcal G_{p,s}$, known as the Grassmann manifold, consists of all
$s$-dimensional linear subspaces of $\mathcal{R}^p$
\citep{Chikuse2012}.  Although $V$ is only identifiable up to right
rotations, the matrix $VV^T$, which defines the plane of variation shared by all
groups, is identifiable.  Later, to emphasize the connection to the
spiked PCA model (\ref{eqn:spiked}), we will write $\Psi_k$ in terms
of its eigendecomposition, $\Psi_k = O_k\Lambda_kO_k$, where $O_k$ are
eigenvectors and $\Lambda_k$ are the eigenvalues of $\Psi_k$ (see
Section \ref{sec:bayes}).

@@

In this section we outline an empirical Bayesian approach for
estimating a low-dimensional shared subspace and the covariance
matrices of the data projected onto this space. As we highlight in
Section \ref{sec:simulation}, if the spiked covariance model holds for
each group individually, then the shared subspace model also
holds because we can take the shared subspace to be the full
$p$-dimensional space. In practice, we can often identify a shared
subspace of dimension $s \ll p$ that preserves most of the variation
in the data.  Our primary objective is to identify the ``best'' shared
subspace of fixed dimension $s < p$.  In Section \ref{sec:em} we
describe an expectation maximization algorithm for estimating the
maximum marginal likelihood of the shared subspace, $VV^T$.  This
approach is computationally tractable for high-dimensional datasets.
Given an inferred subspace, we then seek estimators for the covariance
matrices of the data projected onto this space.  Because seemingly
large differences in the point estimates of covariance matrices across
groups may not actually reflect statistically significant differences,
in Section \ref{sec:bayes} we describe a Gibbs sampler that can be
used to generate estimates of the projected covariance matrices,
$\Psi_k$ and and their associated uncertainty.

% However, in the $n \ll p$ setting, seemingly large
% differences in the point estimates of the covariance matrices across
% groups may not actually reflect statistically significant differences.
% Thus, in Section \ref{sec:bayes} we describe a Gibbs sampler that can
% be used to infer both point estimates and the associated uncertainty
% of the projected covariance matrices, $\Psi_k$.

%TODO: emphasize 1) justification 2) objective of \emph{identifying} this
%subspace when p much greater than n.

@@

We propose an expectation-maximization (EM) algorithm, to estimate the
relevant quantities.  In particular we maximize the marginal
likelihood of $V$, treating $M_k^{-1}$ and $\frac{1}{\sigma_k^2}$ as
the ``missing'' parameters to be imputed.  We assume independent
Jeffreys' prior distributions for both $\frac{1}{\sigma_k^2}$ and
$M_k^{-1}$.  Jeffreys' prior for these quantities corresponds to
$p(\frac{1}{\sigma_k^2}) \propto \sigma_k^2$ and
$p(M_k^{-1}) \propto |M_k^{-1}|^{-(s+1)/2}$.  From the likelihood it
can easily be shown that the conditional posterior for $M_k^{-1}$ is

@@

\noindent which is a gamma($n_k(p-r)/2$,
$\frac{1}{2}tr((I-VV^T)\mathbf{S}_k)$) distribution.  Based on these
results the following EM algorithm can be used:

@@

For the second step (``M-step''), we use a numerical optimization algorithm based on
the Cayley transform to preserve the orthogonality constraints in $V$
\citep{Wen2013}.  Importantly, the complexity of this algorithm is
dominated by dimension of the shared subspace, $s$, not the number of
features $p$.  Thus, our approach is computationally efficient for
relatively small values of $s$, even when $p$ is large.

@@

If $V$ is a basis for a shared subspace, then for each group $k$, most
of the variation in $Y_k$ should be preserved when projecting the data
to this space.  To characterize the extent to which this is true for
different groups, we propose a simple estimator for the proportion of
``signal'' variance that lies on any given subspace.  Specifically,
for an estimated subspace $\hat{V}$, we use the following statistic
for the ratio of the sum of the first $s$ eigenvalues of
$\hat{V}^T\bl \Sigma_k \hat{V}$ to the sum of the first $s$
eigenvalues of $\Sigma_k$:
%
\begin{equation}
\gamma(Y_k: \hat{V}, \hat{\sigma}_k^2) = \frac{||Y_k\hat{V}||_F}{\underset{\widetilde{V} \in \mathcal{V}_{p, s}}{\text{max}}
  ||Y_k\widetilde{V}||_F - \hat{\sigma}_k^2ps/n_k}
\label{eqn:ratio}
\end{equation}
%
\noindent where $||.||_F$ is the Frobenius norm. Our motivation for
this statistic is the fact that if
$\Sigma_k = \hat{V}\Psi_k\hat{V}^T + \sigma_k^2I$ then
$\gamma(Y_k: \hat{V}, \hat{\sigma}_k^2) \approx 1$.  To see this, note
that $\underset{\widetilde{V} \in \mathcal{V}_{p, s}}{\text{max}}\frac{||Y_k\widetilde{V}||_F}{n_k}$
is equivalent to the sum of the first $s$ eigenvalues of the sample
covariance matrix $\bl S_k/n_k$.  With $\hat{\lambda}^{(k)}_i$
the $i$-th eigenvalue of $\bl S_k/n_k$, \citet{Baik2006}
demonstrated that asymptotically as $p, n_k \rightarrow \infty$ with
$p/n_k = \alpha_k$ fixed

\begin{eqnarray}
\nonumber \hat{\lambda}^{(k)}_i &\rightarrow& \lambda^{(k)}_i\left(1 +
                                    \frac{\sigma_k^2\alpha}{\lambda^{(k)}_i
                                    - 1}\right)\\
& \approx& \lambda^{(k)} + \sigma^2_k\frac{p}{n_k}
\end{eqnarray}

Thus,
$\sum_i^s \hat{\lambda}^{(k)}_i \approx \sum_i^s \lambda^{(k)}_i +
\sigma_k^2ps/n_k$.
If we assume that $s$ is fixed in this asymptotic regime, then
$||Y_kV||_F/n_k$ is a consistent estimator for
$\sum_i^s \lambda^{(k)}_i$.  As a consequence, the proposed estimator
will be close to one for all groups $k$ when $\hat{V}\hat{V}^T = VV^T$
and smaller if not.  The metric can provides a useful indicator for
which groups can be reasonably compared on a given subspace and which
groups cannot.  When $\gamma(Y_k: \hat{V}, \hat{\sigma}_k^2)$ is small
for some groups, it may suggest that the rank, $s$, of the inferred
subspace needs to be larger to capture the variation in all groups. To
demonstrate this, in Section \ref{sec:simulation}, we compute these
statistics for inferred subspaces of different dimensions on a single
dataset. In Section \ref{sec:app}, we plot the estimates for subspaces
inferred with real biological data.

%% Problem here! Can't compare different sample sizes??

@@

The EM algorithm presented in the previous section yields point
estimates for $VV^T$ and $\Psi_k$, but does not lead to natural
uncertainty quantification for these estimates.  In this section, we
assume that the subspace $VV^T$ is fixed and known and demonstrate how
we can estimate the posterior distribution for $\Psi_k$.   Note that when
the subspace is known, the posterior distribution of $\Sigma_k$ is
conditionally independent from the other groups, so that we can
independently estimate the conditional posterior distributions for each
group.

% There are many different prior specifications for $\Psi_k$ and
% $\sigma^2_k$ that we could consider for estimating the posterior
% uncertainty.  In this paper, we build on recent interest in the spiked
% covariance model \citep{Donoho2013, Paul2007} and develop tractable
% models for the eigendecomposition of $\Psi_k$.  Such models are
% especially appealing because it is simple to summarize and compare
% $\Psi_k$ across groups in terms their eigenstructure (See Section
% \ref{sec:simulation} and \ref{sec:app}).

Building on recent interest in the spiked covariance model
\citep{Donoho2013, Paul2007} we propose a tractable MCMC algorithm by
specifying priors on the eigenvalues and eigenvectors of $\Psi_k$.  By
modeling the eigenstructure, we can now view each covariance
$\Sigma_k$ in terms of the original spiked principle components model
(\ref{eqn:spiked}).  Equation \ref{eqn:sspsi}, written as a function
$V$, becomes

\begin{align}
\nonumber \Psi_k &= O_k\Lambda_KO_k^T\\
\Sigma_k &= V\Psi_kV^T + \sigma^2_kI
\label{eqn:ss}
\end{align}
%
\noindent Here, we make one further generalization to allow $\Psi_k$
to be a rank $r \leq s$ dimensional covariance matrix on the
$s$-dimensional subspace.  Thus, $\Lambda_k$ is an $r \times r$
diagonal matrix of eigenvalues, and $O_k \in \mathcal{V}_{s,r}$ is the
matrix of eigenvectors of $\Psi_k$.  For any individual group, this
corresponds to the original spiked PCA model (Equation
\ref{eqn:spiked}) with $U_k = VO_k \in \mathcal{V}_{p, r}$.  This
generalization is helpful because it enables us to independently
specify a subspace common to all groups and the possibly lower rank
features on this space that are specific to individual groups.
Although our model is most useful when the covariance matrices are
related across groups, we can use this formulation to specify models
for multiple unrelated spiked covariance models by letting the shared
subspace be full rank, e.g. $s=p$.  We explore this in detail in
Section \ref{sec:simulation}.

@@

\paragraph{Conditional distribution for $\Omega$:} Here we specify the
conditional distribution of the diagonal matrix
$\Omega_k = \Lambda_k(I+\Lambda_k)^{-1} = \text{diag}(\omega_{k1},
... \omega_{kr})$.
We consider a uniform(0,1) prior distribution for each element of
$\Omega$, or equivalently, an $F_{2,2}$ prior distribution for the
elements of $\Lambda$.  The full conditional distribution of an
element $\omega_i$ of $\Omega$ is proportional to the likelihood
function, so

@@

%\subsection{Validation of Inference}
